\documentclass[12pt, letterpaper]{article}

\usepackage{amsmath}
\usepackage{amsfonts}


\title{Fisher-Kolmogorov equations for neurodegenerative diseases}
\author{Andrea Boin, Giacomo Pauletti, Lorenzo Pettenuzzo}
\date{}

\begin{document}
\maketitle
\pagebreak

\tableofcontents
\pagebreak

\section{Introduction}
The objective of this project is to apply various numerical methods to solve the Fisher-Kolmogorov equations to reproduce the results of the paper \cite{diffusion-paper}. The Fisher-Kolmogorov equations can be used to effectively model the spread of misfolded proteins in the brain, a process associated with numerous neurodegenerative diseases.

\subsection{Fisher-Kolmogorov equation}
\[
\begin{cases}
\displaystyle \frac{\partial c}{\partial t} - \nabla \cdot (D \nabla c) - \alpha c(1 - c) = 0 & \text{in } \Omega\\
\displaystyle D \nabla c \cdot \mathbf{n} = 0 & \text{on } \partial \Omega\\
c(t=0)=c_0 & \text{in } \Omega
\end{cases}
\]
\begin{itemize}
    \item $c$: concentration of the misfolded protein in a region of the brain $(0\leq c\leq1)$
    \item $\alpha$: constant of concentration growth
    \item $D$: diffusion coefficient of the misfolded protein.\\
It can be isotropic (a scalar) or anisotropic (a square matrix).\\
In case of anisotropic coefficient the term can be computed as:

$$\underbar{D}=d^{\text{ext}}\underbar{I}+d^{\text{axn}}(\mathbf{n}\otimes\mathbf{n})$$

where $d^{\text{ext}}$ is the extracellular diffusion term, $d^{\text{axn}}$ is the axonal diffusion term and $\mathbf{n}$ the direction of axonal diffusion.

Usually extracellular diffusion is slower than axonal diffusion: $d^{\text{ext}}<d^{\text{axn}}$.

The Fisher-Kolmogorov equation is a \textbf{diffusion-reaction} equation with a nonlinear forcing term that can be used to model population growth. In this case it is used to model the spreading of proteins in the brain.

The interested problem is a \textbf{nonlinear parabolic PDE} with \textbf{Neumann boundary conditions}.
\end{itemize} 
\pagebreak

\subsection{Mesh}
The mesh we used for the simulation is a 3D representation of a hemisphere of the human brain with 21211 points and 42450 cells.\\
To process the mesh with our software, we did convert the format from \textit{.stl} to \textit{.msh} using \textbf{GMSH} with the following procedure:
\begin{enumerate}
    \item Import the mesh (\textit{.stl}) in GMSH
    \item From the left menu, select "geometry $\rightarrow$ add $\rightarrow$ volume"
    \item Save the new generated \textit{.geo} file
    \item Define the 3D mesh: "mesh $\rightarrow$ define $\rightarrow$ 3D"
    \item Export the file as \textit{.msh}: "file $\rightarrow$ export $\rightarrow$ msh"
\end{enumerate}

\noindent The mesh isn't associated to any function that could provide information about axonal orientation so an anisotropic model can't be used to accurately simulate the evolution of the system.
\noindent Also there is no distinction between white and grey matter in the mesh. Different kinds of matter have different reaction coefficients so we used an average of the two for our simulations. 

\subsection{Weak formulation and semi-discretized formulation}
By choosing a domain $V=H^1=\{v\in L^2|\nabla v\in L^2\}$ and considering a time domain $(0, T)$, the weak formulation of the problem is:

\vspace{1em}
\noindent Find $c(t)\in V$ such that $\forall v\in V$ and $\forall t\in(0,T)$:
$$\begin{cases}\int_\Omega\frac{\delta c}{\delta t}vd\Omega+\int_\Omega D\nabla c\nabla vd\Omega-\int_\Omega\alpha c(1-c)vd\Omega=0\\c(t=0)=c_0\end{cases}$$

\vspace{1em}
\noindent By renaming:
\begin{itemize}
    \item $a(c,v)=\int_\Omega D\nabla c\nabla vd\Omega$
    \item $n(c,v)=-\int_\Omega\alpha c(1-c)vd\Omega$
\end{itemize}

\noindent By introducing a triangulation $T_h=\{K|\Omega=\bigcup K\}$ of the domain $\Omega$ and defining with it a polynomial space $$X_h=\{v_h\in C^0(\bar\Omega)|v_{h|k}\in\mathbb{P}^r(K),\forall K\in T_h\}$$ we can obtain the discrete space $V_h=V\cap X_h$ for our discrete formulation.

\noindent The semi-discrete formulation can then be written as:

\vspace{1em}
\noindent
Find $c_h\in V_h$ such that, $\forall v_h\in V_h$ and $\forall t\in(0,T)$:
$$\int_\Omega\frac{\delta c_h}{\delta t}v_hd\Omega+a(c_h,v_h)+n(c_h,v_h)=0$$
$$c_h(t=0)=c_{h,0}$$

\section{Methods}

We studied the problem with 3 methods and implemented 2 of them algorithmically:
\begin{itemize}
    \item An \textbf{explicit} scheme in which all terms have been treated explicitly to handle the nonlinear part of the model.
    \item A \textbf{mixed explicit/implicit} scheme in which the linear terms have been treated implicitly while the nonlinear terms explicitly to get rid of nonlinarities.
    \item An \textbf{implicit} scheme in which all terms in the equation have been treated implicitly, and then the nonlinear parts have been solved with the Newton method.
\end{itemize}

\noindent To obtain a full discretization of the problem we need to partition the time domain in $N$ partitions of size $\Delta t$, obtaining $(0, T)=(0,N\Delta t)=\bigcup_{n=1}^N(t^n, t^{n+1}]$ where $t^{n+1}-t^n=\Delta t$, $t^0=0$ and $t^N=T$. We can then use an upper-index notation to identify time dependent elements: $c^n = c(t^n)$.

\subsection{Explicit scheme}

The fully discrete formulation for the \textbf{explicit} scheme becomes:

\vspace{1em}
\noindent
Find $c_h(t)\in V_h$ such that, $\forall v_h\in V_h$, $c_h^0=c_{h,0}$ and $\forall n\in\{0, N\}$:
$$\int_\Omega\frac{c_h^{n+1}-c_h^n}{\Delta t}v_hd\Omega+a(c_h^n,v_h)+n(c_h^n,v_h)=0$$

\noindent By introducing a basis $\{\phi_i\}$ for the space $V_h$ the problem can be written as:

\vspace{1em}
\noindent
Find $c_h(t)\in V_h$ such that $c_h^0=c_{h,0}$ and $\forall n\in\{0, N\}$:
$$Mc^{n+1}=F^n$$
where the \textbf{mass matrix} can be computed as:
$$M_{ij}=\frac1{\Delta t}\langle\phi_j,\phi_i\rangle$$
and the \textbf{forcing term} is:
$$F_i^n=\frac1{\Delta t}\langle c_{h,j}^n,\phi_i\rangle-a(c_{h,j}^n,\phi_i)-n(c_{h,j}^n,\phi_i)$$

\subsubsection{Stability and Accuracy}
The accuracy is $O(\Delta t)$ for time and $O(h^2)$ for space.

% TODO: check values correctness
\noindent The stability condition of the explicit scheme is: $\Delta t\leq\min(\frac{h^2}{2D}, \frac2\alpha)$. The first term acts as a bottleneck for the method. With our values for example ($h=1[cm], D=1.5[cm/year], \alpha=0.5[1/year]$), $\Delta t\leq\min(\frac13, 4)=\frac13$. The following methods allow for a larger choice of $\Delta t$ and are generally faster so we decided to implement them.

\subsection{Mixed explicit/implicit scheme}
The full discretization for the \textbf{mixed explicit/implicit} scheme is:

\vspace{1em}
\noindent
Find $c_h\in V_h$ such that, $\forall v_h\in V_h$ and $c_h(t=0)=c_{h,0}$:
$$\int_\Omega\frac{c_h^{n+1}-c_h^n}{\Delta t}v_hd\Omega+a(c_h^{n+1},v_h)+n(c_h^n,v_h)=f(v_h)$$

\noindent The problem can be rewritten, by introducing a basis $\{\phi_i\}$ for $V_h$ as:

\vspace{1em}
\noindent
Find $c_h(t)\in V_h$ such that $c_h^0=c_{h,0}$ and $\forall n\in\{0, N\}$:
$$Mc^{n+1}=F^n$$
where the \textbf{mass matrix} can be computed as:
$$M_{ij}=\frac1{\Delta t}\langle\phi_j,\phi_i\rangle+a(\phi_j,\phi_i)$$
and the \textbf{forcing term} is:
$$F_i^n=\frac1{\Delta t}\langle c_{h,j}^n,\phi_i\rangle-n(c_{h,j}^n,\phi_i)$$

\subsubsection{Algorithm}


\subsubsection{Stability and Accuracy}
The accuracy for this method is the same as the explicit one: $O(\Delta t)$ for time and $O(h^2)$ for space.

\noindent The stability condition though is better: $\Delta t\leq\frac2\alpha=4$ allowing for a larger choice of $\Delta t$ and a quicker convergence.

\subsection{Implicit scheme}
\subsubsection{Stability and Accuracy}

Stability and accuracy analysis of the implemented methods.

\section{Results and algorithmic comparation}
Results of the project for all the implemented algorithms and graphs.
% TODO: graph about explosion of disease confirming real evidence
% TODO: say some proteins might act at different percentages of concentration
% TODO: parallelism

\begin{thebibliography}{9}
    \bibitem{diffusion-paper}
    J. Weickenmeier, M. Jucker, A. Goriely, and E. Kuhl. A physics-based model explains the prion-like features of neurodegeneration in Alzheimer’s disease, Parkinson’s disease, and amyotrophic lateral sclerosis. Journal of the Mechanics and
    Physics of Solids, 124:264–281, 2019.
\end{thebibliography}

\end{document}